Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-22 Paper/Paper.tex
changed tex

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{ambient assisted living, capacitive proximity sensors, activity recognition, user context}
\end{abstract}

@@

[ AAL scenario as motivation for research ]

Sensing a persons activity is being researched for several years now, raising even more interest in it by the advanced technology of mobile phones, which include multiple sensors and therefore enable activity recognition with such a device \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005} to identify the activities performed. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify fine-grained activities just by using a single modality like the accelerometer.

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show how an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
% For this we obtained a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and integrate a capacitive proximity sensor unobtrusively into the wristband.
For this we obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrate a capacitive proximity sensor unobtrusively into the wristband.

There are several intuitive examples for which a combination of accelerometer-based activity recognition with a capacitive proximity sensor reveals its strength. For example, it may be conducted which material is placed underneath the hand. The capacitive proximity sensor would return a different measurement result for a hand placed on a couch covered with fabric than a hand placed a wooden table. Moreover, the approximate distance to objects can be exploited to identify activities like grasping into a locker or a refrigerator to prepare food.

A wristband worn capacitive proximity sensor requires a shield that eliminates the influence of the grounded arm directly underneath the sensor. Using this setup, we can detect the proximity to a grounded object in the environments for distances up to 20 cm. Especially for mobile devices, it is required that the sensor only draws a very small amount of power. Thus, other proximity sensing input modalities like ultrasound or optical measurements are not applicable for this type of mobile application.

@@

The remainder of this paper is  is structured as follows: First, Section \ref{sect:related} is dedicated to related work work in activity recognition, in particular considering approaches with multiple sensor modalities. Section \ref{sect:hardware} will present the hardware based on which a wrist-worn sensor prototype was built, fusing together an inertial data logger and a capacitive sensor integrated in a wristband. The experimental setup, the scripted scenario with daily activities and the activity recognition evaluation results, showing the performance boost of the capacitive sensor unit, are given in Section \ref{sect:experiment}. The paper is wrapped up with a conclusions section enumerating the findings of the paper, as well as pointing out future research potential.

@@

Activity recognition research relying on wearable sensors mostly considers inertial data from the participants body to infer performed activities, such as in the works of \cite{Ravi2005,Bao2004,Srinivasan2010,Amft2005}. The acceleration data is often augmented with data from sensors such as gyroscopes \cite{Holleczek_2010}, magnetometers \cite{Altun_2010}, ambient light \cite{Borazio2012} or ambient and skin temperature \cite{Krause_2003}, aiming at extracting more motion and environmental context of the user. Researchers in \cite{wyss2010recognition}, for example, use heart rate information additionally to the accelerometer to detect military related activities like lifting and lowering loads or even digging. In \cite{Ward_2006}, workshop assembly activities were detected by augmenting acceleration sensors with microphones.

The works of \cite{Fishkin_2005} and \cite{patterson2005fga} show that detecting objects the user touches or uses can be very helpful for activity recognition. By using RFID readers that are embedded in gloves or bracelets at the wrist and RFID tags attached to various objects of interest, one can detect the object grasped and used by the user, thus aiding the activity recognition in various application scenarios, such as activities of daily living \cite{phealth:maja} and \cite{Philipose_2004}, activity tracking in car manufacturing \cite{Stiefmeier08}, or household and gardening activities \cite{berlin_laerhoven_tei_2010}.

Our approach to enhance the inertial data from a wearable sensor is quite similar to the RFID scenarios just mentioned, as it also relies on a single wearable sensor and thus is aiming at an unobtrusive deployment. The main difference lies in the fact that we do not need accurate detection of tagged objects, but are only interested in the proximity to various objects which, as we expect, could boost activity recognition.

[ check the wording ! ]

@@

In the field of capacitive proximity sensing, three different measurement modes shown in Figure \ref{fig:modes} were identified by Smith et al. \cite{Smith1999}: transmit mode, shunt mode and loading mode. Transmit mode is based on a varying electric potential coupled to an object that can be measured by a capacitive proximity sensor next to that object. Shunt mode applies two electrodes, a transmit and a receive electrode, that can measure capacitance changes produced by objects disturbing the electric field between the two electrodes. In loading mode, a single electrode builds up an electric field to any grounded object in the environment. By measuring the capacitance, conclusions can be made upon the proximity and nature of an object. In our work we apply loading mode since it requires only a single electrode that can be integrated invisibly into the wristband.

@@

In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. For example, Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

\section{Hardware}
\label{sect:hardware}
This section presents the two components of the required hardware, the wrist-worn activity data logger tailored to capture acceleration data, and the capacitive proximity sensor used for distance measurements.

\subsection{Activity Data Logger}
The HedgeHog sensor \cite{hedgehog} is a custom designed and built wearable data logger aiming at long-term deployments in activity recognition scenarios. Due to its small form-factor (37x32x16mm) and weight, this sensor, being worn at the wrist, is an unobtrusive way to record relevant motion data.

@@

The 3D accelerometer sensor is being sampled at $100Hz$, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card. The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor, which is described in detail in the next section.

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the arm that is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active. This low power consumption qualifies the sensor for wearable proximity sensing applications.

@@

The experiment setup aims to depict a typical scenario in daily life. Especially in the field of Ambient Assisted Living (AAL), it is desired to monitor activities like drinking, preparing lunch and sleeping. A fine-grained monitoring of such activities may help elderly or people suffering from mental diseases to maintain a healthy day/night rhythm and take action if irregularities occur. Figure \ref{fig:cap_sensor} shows the modified HedgeHog activity logger that has been extended with a capacitive proximity sensor. The wristband has two electrodes, a sensing electrode underneath a slightly bigger shield electrode.

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, [kitchen preparations!] taking plate, cutlery, bread, marmalade from a shelf or a locker, sitting at the table, preparing a bread with marmalade, eating the bread, drinking water, walking and sleeping. Some of those activities are very hard to recognize without any further contextual information, when the data is limited to a single modality like a 3D accelerometer. For example, sitting at the table and sitting on a couch are very similar activities. We aim to show that the data basis can be significantly improved by the additional input modality.

In order to evaluate if capacitive proximity sensors in wrist-bands can enhance the performance of activity recognition, we have conducted an evaluation with 7 test persons. All test persons received a basic plot with the activities they were supposed to perform. They were not given any instructions about the way they are supposed to perform the activities. After manual labeling, we used this test-set as ground truth and performed a 4-fold cross-validation on an support-vector machine (SVM) classifier on each user, once with and once without including the data of the capacitive proximity sensor into the feature set. We used an SVM classifier because of its high relevance in activity recognition and its fast performance.

@@

The classifier was trained with basic features that were extracted from a sliding window of 1 second width. Our first tests have showed that greater window sizes do not provide better results. In order to suppress noise contained in the capacitive proximity sensing data, we applied a moving average filter with a kernel size of 10. The final feature set contained the arithmetic mean, min, max, median and standard variance for each accelerometer axis and the capacitive proximity signal. These simple feature types represent standard features applied in activity recognition.

>>>>

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Persons affected from physical and mental restrictions and their care givers can profit significantly from unobtrusive activity monitoring solutions. For example, formal care givers could equip a person with a wearable activity monitoring solution that evaluates the course of a disease or the influence of an adapted medication. This scenario is especially relevant for persons suffering from dementia who have limitations in organizing their daily activities. A simple wearable activity monitoring solution could analyze activities performed in daily life, such as drinking, eating and sleeping habits.

Sensing a person's activity is an active research topic with a raising interest due to the advancement in mobile phone technology. These devices include multiple sensors and therefore enable the recognition of daily activities \cite{brezmes2009activity}. Current wearable activity recognition systems are able to unobtrusively capture and recognize a person's activities throughout the whole day. These systems often rely on inertial sensor data that is captured by wearable sensors embedded in a mobile device \cite{brezmes2009activity} or attached to the body \cite{Ravi2005}. Usually, single sensor modalities are used or duplicated to detect the activities. However, it is a great challenge to identify fine-grained activities just by using a single modality like the accelerometer.

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
% For this we obtained a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and integrate a capacitive proximity sensor unobtrusively into the wristband.
We obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

There are several intuitive examples for which a combination of accelerometer-based activity recognition with a capacitive proximity sensor reveals its strength. For example, it may be conducted which material is placed underneath the wristband. The capacitive proximity sensor would return a different measurement result for a hand placed on a couch covered with fabric than a hand placed a wooden table. Moreover, the approximate distance from the wristband to objects can be exploited to identify activities like grasping into a locker or a refrigerator to prepare food.

A wrist-worn capacitive proximity sensor requires a shield that eliminates the influence of the grounded arm directly underneath the sensor. Using this setup, we can detect the proximity to a grounded object in the environments for distances up to 20 cm. Especially for mobile devices, it is required that the sensor only draws a very small amount of power. Thus, other proximity sensing input modalities like ultrasound or optical measurements are not applicable for this type of mobile application.

@@

The remainder of this paper is structured as follows: First, Section \ref{sect:related} is dedicated to related work work in activity recognition, in particular considering approaches with multiple sensing modalities. Section \ref{sect:hardware} presents the hardware based on which a wrist-worn sensor prototype was built, fusing an inertial data logger and a capacitive sensor integrated into a wristband. The experimental setup, the scenario with daily activities and the activity recognition evaluation results, showing the performance boost of the capacitive sensor unit, are given in Section \ref{sect:experiment}. The paper is wrapped up with a conclusion enumerating the findings of the evaluation, as well as pointing out future research potential.

@@

Activity recognition research relying on wearable sensors mostly considers inertial data from the participants body to infer performed activities, such as in the works of \cite{Ravi2005,Bao2004,Srinivasan2010,Amft2005}. The acceleration data is often augmented with data from sensors such as gyroscopes \cite{Holleczek_2010}, magnetometers \cite{Altun_2010}, ambient light \cite{Borazio2012} or ambient and skin temperature \cite{Krause_2003}, aiming to extract a more detailed environmental user context. Researchers in \cite{wyss2010recognition} use heart rate information as an addition to the accelerometer data to detect military related activities like lifting and lowering loads or even digging. In \cite{Ward_2006}, workshop assembly activities are detected by augmenting acceleration sensors with microphones.

The works of \cite{Fishkin_2005} and \cite{patterson2005fga} show that detecting touched and used objects can be very helpful for activity recognition. By using RFID readers that are embedded in gloves or bracelets at the wrist and RFID tags attached to various objects of interest, one can detect the object grasped and used by the user, thus aiding the activity recognition in various application scenarios, such as activities of daily living \cite{phealth:maja} and \cite{Philipose_2004}, activity tracking in car manufacturing \cite{Stiefmeier08}, or household and gardening activities \cite{berlin_laerhoven_tei_2010}.

Our approach to enhance the inertial data from a wearable sensor is comparable to the RFID scenarios just mentioned, as it also relies on a single wearable sensor and an unobtrusive deployment. The main difference lies in the fact that we do not consider an accurate detection of tagged objects, but the proximity to various objects and the environment.

@@

In the field of capacitive proximity sensing, three different measurement modes (shown in Figure \ref{fig:modes}) were identified by Smith et al. \cite{Smith1999}: transmit mode, shunt mode and loading mode. Transmit mode is based on a varying electric potential coupled to an object that can be measured by a capacitive proximity sensor next to that object. Shunt mode applies two electrodes, a transmit and a receive electrode, that can measure capacitance changes produced by objects disturbing the electric field between the two electrodes. In loading mode, a single electrode builds up an electric field to any grounded object in the environment. By measuring the capacitance, conclusions can be made upon the proximity and nature of an object. In our work we apply loading mode since it requires only a single electrode that can be integrated invisibly into the wristband.

@@

In previous works, capacitive proximity sensors were applied in various fields of human-computer-interaction. Wimmer et al. and Grosse-Puppendahl et al. presented gesture recognition systems \cite{Wimmer,Grosse-puppendahl2012} as well as smart furniture that can sense human activities \cite{Wimmer} and classify human postures \cite{Grosse-puppendahl2011}. Cheng et al. have investigated the possibility of using capacitive sensors for activity recognition by measuring shape changes of muscles and skin \cite{Cheng2010}. To our knowledge, capacitive proximity sensors have not been embedded into a wearable device to enhance the performance of activity recognition.

\section{Hardware}
\label{sect:hardware}

This section presents the two components of the required hardware, the wrist-worn activity data logger tailored to capture acceleration data, and the capacitive proximity sensor used for distance measurements.

\subsection{Activity Data Logger}

The HedgeHog sensor \cite{hedgehog} is a custom designed wearable data logger aiming at long-term deployments in activity recognition scenarios. Due to its small form-factor (37x32x16mm) and weight, this wrist-worn sensor is an unobtrusive way to record relevant motion data.

@@

The 3D accelerometer sensor is being sampled at $100Hz$, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card. The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor that is described in detail in the next section.

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the underlying skin. Without shielding no meaningful measurements could be performed, since the arm's skin is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.

@@

The experiment setup aims to depict a typical scenario of person in daily life. Especially in the field of Ambient Assisted Living (AAL), it is desired to monitor activities like drinking, preparing lunch and sleeping. A fine-grained monitoring of such activities may help elderly or people suffering from mental diseases to maintain a healthy day/night rhythm and take action if irregularities occur. Figure \ref{fig:cap_sensor} shows the modified HedgeHog activity logger that has been extended with a capacitive proximity sensor. The wristband has two electrodes, a sensing electrode underneath a slightly bigger shield electrode.

The recorded test set contains the following activities: opening door, sitting on a couch, lying on a couch, putting kitchen equipments from a shelf and out of a locker, preparing a bread with marmalade, eating the bread, pouring and drinking water, walking and sleeping. The relations of these activities to environmental objects are given in Table \ref{tab:activities}. Some of those activities are very hard to recognize when the data is limited to a single modality like a 3D accelerometer. For example, sitting at the table and sitting on a couch are very similar activities. We aim to show that the data basis can be significantly improved by the additional input modality.

@@

In order to evaluate if capacitive proximity sensors in wrist-bands can enhance the performance of activity recognition, we have conducted an evaluation with 7 test persons. All test persons received a basic plot with the activities they were supposed to perform. They were not given any instructions about the way they are supposed to perform the activities. After manual labeling, we used this test-set as ground truth and performed a 4-fold cross-validation on an support-vector machine (SVM) classifier on each user. The cross-validation was performed once with and once without including the data of the capacitive proximity sensor into the feature set. We chose an SVM classifier because of its high relevance in activity recognition and its fast performance.

The classifier was trained with basic features that were extracted from a sliding window of 1 second width. Our first tests have shown that greater window sizes do not provide better classification results. In order to suppress noise contained in the capacitive proximity sensing data, we applied a moving average filter with a kernel size of 10. The final feature set contained the arithmetic mean, min, max, median and standard variance for each accelerometer axis and the capacitive proximity signal. These simple feature types represent standard features applied in activity recognition. Since we aim to show an improvement using the new modality, the selection of features and classifiers does not represent the primary focus of this paper.
